Why Did Russia Fare So Poorly during COVID: An Analysis of Why One of the Best Prepared Countries Could Not Control Excess Deaths

### 2024-02-13 paper/paper.qmd
Abstract and Measurement
- Added Abstract
- Added measurement sections.

---
title: "Why Did Russia Fare So Poorly during COVID: An Analysis of Why One of the Best Prepared Countries Could Not Control Excess Deaths"
author: 
  - Aryaman Sharma
  - Aviral Bhardwaj
  - Janel Gilani
thanks: "Code and data are available at: https://github.com/Ary4m3n/covid-effect-russia.git; Replication on Social Science Reproduction platform available at: https://www.socialsciencereproduction.org/reproductions/8938b287-6681-4a5e-8993-42011a51be58/index"
date: today
date-format: long
abstract: "GHS Index"
format: pdf
toc: true
number-sections: true
bibliography: references.bib
---

@@

When discussing COVID-19, (1) issues with underreported or missing data, particularly related to non-responses, are prevalent, (2) the lack of a standardized definition of COVID-19 deaths, which can lead to measurement errors (3) vastness of countries poses logistical challenges in reporting and conducting surveys, making it difficult to gather comprehensive data (4) variations in testing methodologies and availability of resources across regions can further distort the data.

Many countries, especially developing nations like India, Brazil, and Russia, poses above issue. This poses a significant challenge in accurately measuring the pandemic's impact on a country. For example, the excess death rate, a variable in our analysis, measures the number of deaths occurring in a specific period above the expected number. This metric helps estimate the pandemic's impact. However, due to data truncation, the actual death rate isn't always precise, leading to inaccuracies in calculating the excess death rate. This introduces measurement errors, undermining the accuracy of both the actual death rate and the excess death rate.

With respect to measurement and accounting for the GHS index, it encompasses various measurements grouped into six broad categories: (1) prevention, (2) detection and reporting, (3) rapid response, (4) health system, (5) compliance with international norms, and (6) risk environment. However, various measurement errors and qualitative biases can arise from the subjective nature of these categories. For example, the "prevention" category includes measurements such as "vaccination coverage" and "biosecurity and biosafety." However, the definition of "biosecurity and biosafety" can vary across countries, leading to measurement errors. Additionally, the GHS index is based on self-reported data, which can be subject to biases and inaccuracies. This can lead to measurement errors in the GHS index, which can further distort the analysis.

@@

>>>>

---
title: "Why Did Russia Fare So Poorly during COVID: An Analysis of Why One of the Best Prepared Countries Could Not Control Excess Deaths"
author: 
  - Aryaman Sharma
  - Aviral Bhardwaj
  - Janel Gilani
thanks: "Code and data are available at: https://github.com/Ary4m3n/covid-effect-russia.git; Replication on Social Science Reproduction platform available at: https://www.socialsciencereproduction.org/reproductions/8938b287-6681-4a5e-8993-42011a51be58/index"
date: today
date-format: long
abstract: " Despite Russia being deemed well-prepared for pandemics, it faced an unprecedented rise in death rates compared to other nations. This study aims to replicate Nuzzo and Ledesma's findings, examining COVID-19's impact on Russia while considering variables such as the Global Health Security (GHS) Index and excess death rates. Our analysis revealed a significantly elevated excess death rate in Russia, contrasting with the lower rates in more prepared countries. This research sheds light on the COVID-19 outbreak's impact on Russia and identifies areas for improvement for well-prepared countries to mitigate future pandemics' effects, emphasizing the importance of better planning and regulations."
format: pdf
toc: true
number-sections: true
bibliography: references.bib
---

@@

With respect to measurement and accounting for the GHS index, it encompasses various measurements grouped into six broad categories: (1) prevention, (2) detection and reporting, (3) rapid response, (4) health system, (5) compliance with international norms, and (6) risk environment. These values are calculated based on the data available for each country which are self-reported and are then used to calculate the overall GHS index [@GHSIndexMethodology].`

In addition to tracking deaths, a crucial data point during the COVID-19 pandemic is the excess death rate, representing the number of deaths occurring within a specific period above the expected baseline. This metric serves as a key indicator of the pandemic's impact on mortality. In our dataset, death rates are standardized to a per 100,000 population basis, facilitating meaningful comparisons across countries.

The process of collecting this data is primarily overseen by institutions like the Institute for Health Metrics and Evaluation (IHME), which collates information from various sources available for each country. Deaths included in the excess death rate calculation encompass those directly attributed to COVID-19, as well as those indirectly influenced by the pandemic where COVID-19 appeared on the death certificate. For instance, deaths resulting from delayed treatment due to healthcare system strain caused by the pandemic are also accounted for.

Observation of deaths involves various mechanisms. Initially, healthcare facilities and authorities report COVID-19-related deaths through diagnostic tests and medical records. Deaths with an underlying cause of death of COVID-19 are not included in these estimates. For example, in some cases, COVID-19 may have contributed to the death, but the underlying cause of death was another cause, such as terminal cancer. Additionally, mortality statistics are compiled from death certificates and official registries. Tallying these deaths are then put into the Government's official records.

Once observed deaths are verified and categorized, they are tallied according to established criteria for inclusion as defined by IHME. the excess death rate calculation. This meticulous approach ensures the accuracy and reliability of the data used to compute the excess death rate for each country.

However, due to data truncation, the actual death rate isn't always precise, leading to inaccuracies in calculating the excess death rate. This introduces measurement errors, undermining the accuracy of both the actual death rate and the excess death rate. This is a significant challenge in accurately measuring the pandemic's impact on a country.

<!-- When discussing COVID-19, (1) issues with underreported or missing data, particularly related to non-responses, are prevalent, (2) the lack of a standardized definition of COVID-19 deaths, which can lead to measurement errors (3) vastness of countries poses logistical challenges in reporting and conducting surveys, making it difficult to gather comprehensive data (4) variations in testing methodologies and availability of resources across regions can further distort the data. -->

<!-- Many countries, especially developing nations like India, Brazil, and Russia, poses above issue. This poses a significant challenge in accurately measuring the pandemic's impact on a country. For example, the excess death rate, a variable in our analysis, measures the number of deaths occurring in a specific period above the expected number. This metric helps estimate the pandemic's impact. However, due to data truncation, the actual death rate isn't always precise, leading to inaccuracies in calculating the excess death rate. This introduces measurement errors, undermining the accuracy of both the actual death rate and the excess death rate. -->

<!-- With respect to measurement and accounting for the GHS index, it encompasses various measurements grouped into six broad categories: (1) prevention, (2) detection and reporting, (3) rapid response, (4) health system, (5) compliance with international norms, and (6) risk environment. However, various measurement errors and qualitative biases can arise from the subjective nature of these categories. For example, the "prevention" category includes measurements such as "vaccination coverage" and "biosecurity and biosafety." However, the definition of "biosecurity and biosafety" can vary across countries, leading to measurement errors. Additionally, the GHS index is based on self-reported data, which can be subject to biases and inaccuracies. This can lead to measurement errors in the GHS index, which can further distort the analysis. -->

@@

It should be noted that different countries employ varying methods of reporting their data, including mortality rates and excess death rates. Therefore, this study primarily relied on observational data, but it's crucial to account for differential biases in the data. For instance, some countries might underreport their deaths, while others might overreport them. Such discrepancies directly impact the analysis conducted in this paper.

Furthermore, COVID-19 policies were predominantly implemented at the state and local levels, introducing variability throughout the pandemic. Differences in policy implementation among these communities, coupled with the presence of unmeasured differences, complicate the ability to draw causal inferences.
